---
title: "Sev Plant Summary"
author: "Tim Farkas"
date: "4/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##### Modify Trait Data For Analysis and Merge with CV Data
```{r}
raw_traits <- read_csv("~/Dropbox/Projects/Sev/Plants/Data/All Traits/sev_all_traits_Nov2018.csv") %>%
  filter(! kartez %in% c("DAJA?")) %>%
  mutate_at(vars(kartez), funs(ifelse(. == "ARPU9", "ARIST", .))) %>%
  mutate_at(vars(kartez), funs(ifelse(. %in% c("SPCO4", "SPFL2"), "SPORO", .))) %>%
  mutate_at(vars(kartez), funs(ifelse(. %in% c("ASFE2", "ASNU4"), "ASTRA", .))) %>%
  mutate_at(vars(a_p), funs(ifelse(. == "a", 0, 1))) %>%
  group_by(kartez) 

load(file="~/Dropbox/Projects/Sev/Plants/Data/Biomass/sev_plant_CV.RData")

all_traits <- 
  inner_join(raw_traits %>% summarize_at(vars(family:apgf, -a_p), funs(max)),
             raw_traits %>% summarize_at(vars(a_p, avg.height_A:seed_mass), funs(mean)), 
             by = "kartez") %>%
  inner_join(temp_cv, by="kartez") %>%
  mutate_at(vars(lma), funs(ifelse(. > 0.1, NA, .))) %>%
  mutate_at(vars(p95.height_A), funs(ifelse(. > 100, NA, .))) %>%
  mutate_at(vars(sdens), funs(ifelse(. > 0.003, NA, .))) %>%
mutate_at(vars(rdens), funs(ifelse(. > 0.003, NA, .))) %>%
  mutate_at(vars(fr.diam), funs(ifelse(. > 0.15, NA, .))) 


```

##### Basic traits exploration
```{r}

ggplot(all_traits) + 
  geom_histogram(aes(x=seed_mass))

ad2 <- ad_ag %>%
  select(a_p, p95.height_A, ldmc, lma, area, cn, sdmc:rdens) %>%
  mutate_at(vars(a_p), as.numeric) %>%
  filter(p95.height_A < 150, lma < .09, rdens < .002, cn < 60)

ggplot(data=ad2, aes(ad2$cn)) + geom_histogram()
filter(ad2, p95.height_A > 100)

ggpairs(ad2)
```

#### PICs

pics() function takes x = a data frame, cols = a character vector of column names, phy = a tree file, center = logical to mean center all PICs, and a print.tree = logical whether to print the phylogeny. pics() returns a data frame with PICs for each variable, and automatically includes only complete cases across cols. 

```{r pics() Function}

pics.old <- function(x, cols=NULL, phy=tree158, taxa="taxon", center=TRUE, print.tree=FALSE)  {
  
  require(ape)
  x <- as.data.frame(x)
  x <- x[!is.na(x[[taxa]]), ]
  tip.names <- x[[taxa]]
  
  phy$node.label <- paste("node", 1:phy$Nnode)
  
  if(is.null(cols)) dd <- x[ , colnames(x) != taxa, drop=FALSE] else dd <- x[ , cols[cols != taxa], drop=FALSE] 
  
  rownames(dd) <- tip.names
  dd <- dd[complete.cases(dd), , drop=FALSE] # get only data with mutually complete records
  
  # prune data and tree
  dd <- dd[rownames(dd) %in% intersect(rownames(dd), phy$tip.label), , drop=FALSE] # prune data
  pruned <- drop.tip(phy, setdiff(phy$tip.label, rownames(dd))) # prune tree
  if(!setequal(pruned$tip.label, rownames(dd))) # check that taxa match
    return("ERROR: tree and trait taxa do not match")
  pruned <- multi2di(pruned, random=FALSE) # resolve multifurcations with branches of zero length
  # pruned$node.label <- 1:pruned$Nnode # add node labels
  if(print.tree) {
    plot.phylo(pruned, show.node.label = TRUE, cex=.5, adj=0, 
               use.edge.length = FALSE)
  }
  # pruned$node.label <- paste("node", 1:pruned$Nnode) # add better node labels
  
  picmat <- as.data.frame(sapply(dd, function(y) pic(y, pruned))) # make PIC data for all columns
  if(center) picmat <- as.data.frame(t(t(picmat) - colMeans(picmat))) # mean center
  return(picmat)
  
}



```

```{r Better pics() function}
pics <- function(x, # a tibble or data frame with trait data
                 cols=NULL, # a character string of trait columns
                 phy, # a phylogenetic tree of class "phylo" & order "cladewise"
                 taxa, # a character string of the taxon column, to match tips in phy
                 center=TRUE, 
                 scale=TRUE) { # logical indicator whether to mean center PIC output
  
  require(ape)
  require(dplyr)
  require(geiger)
  

  # format data frame input
  dd <- x %>% 
    as.tibble() %>% # coerce to tibble
    rename(taxon = taxa) %>% # identify taxon column
    filter(!is.na(taxon)) # remove NA taxa
  
  # specify all columns if none given by user
  if(is.null(cols)) cols <- names(dd %>% select(-taxon)) 
  
  # prune tree for matches only
  phy <- drop.tip(phy, setdiff(phy$tip.label, dd$taxon))
  phy <- multi2di(phy, random=FALSE) 
  phy$node.label <- paste("node", 
                          formatC(c(1:phy$Nnode), 
                                  digits=0, width=nchar(phy$Nnode), flag="0"))
  
  # get single PICs function
  pic1 <- function(col  # 2 column data frame or tibble with first column taxon, 
                   # second column trait data, possibly with NAs
                   ,  phy
                   ,  scale=scale) {  # a .tre object to patch first column of col

    # remove NA and add taxon as rownames for apre::pic 
    col <- col %>% 
      filter(!is.na(.[2])) %>%
      as.data.frame()
    rownames(col) <- col$taxon
    col <- col %>% 
      select(2)
    
    # prune data and tree
    col <- col[rownames(col) %in% intersect(rownames(col), phy$tip.label), 
               , drop=FALSE] 
    pruned <- drop.tip(phy, setdiff(phy$tip.label, rownames(col)))
    
    # resolve multifurcations with branches of zero length
    # pruned <- multi2di(pruned, random=FALSE) 
    
    # get pic column
    pic.out <- pic(col[[1]], pruned, scaled=scale
                   ,var.contrasts=TRUE
                   ) #%>%
    #   tibble(names(.)) %>%
    #   select(2, 1)
    # names(pic.out) <- c("node", names(col))
    
    return(pic.out)
    
  }  
  
  # generate list of pics for all traits 
  pic_list <- lapply(cols, function(y) {
    
    trait <- data.frame(taxon=dd[["taxon"]], dd[[y]])
    names(trait)[2] <- y
    as.data.frame(pic1(col=trait, phy=phy, scale=scale))
    
  })                     
  
  ojoin_list <- function(right) { # recursive joining function
    
    left <- right[1]
    right <- right[-1]
    
    if(length(right) == 0) return(left[[1]])
    print("continued")
    if(length(right) == 1) return(full_join(left[[1]], right[[1]], by="node"))
    
    full_join(left[[1]], ojoin_list(right), by="node")
    
  }
  
  pics_only <- lapply(pic_list, 
                      function(x) data.frame(node=rownames(x),
                                                       contrasts=x$contrasts))
  vars_only <- lapply(pic_list, 
                      function(x) data.frame(node=rownames(x), 
                                                       variance=x$variance))
  
  # join list of pics into single data frame, joining on node
  
  if(center) {pics_df <- suppressWarnings(ojoin_list(pics_only)) %>% 
    mutate_at(vars(2:ncol(.)), funs(. - mean(., na.rm=TRUE)))} else
      pics_df <- ojoin_list(pic_list)
  
  vars_df <- suppressWarnings(ojoin_list(vars_only))
  
  names(pics_df) <- c("node", cols)
  names(vars_df) <- c("node", cols)
  
  outlist <- list(pics_df=pics_df[order(pics_df$node), ], 
                  vars_df=vars_df[order(vars_df$node), ])
  
  return(outlist)
  
}
```

```{r Better pics() function with transformation}
pics <- function(x, # a tibble or data frame with trait data
                 cols=NULL, # a character string of trait columns
                 phy, # a phylogenetic tree of class "phylo" & order "cladewise"
                 taxa, # a character string of the taxon column, to match tips in phy
                 center=TRUE, 
                 scale=TRUE) { # logical indicator whether to mean center PIC output
  
  require(ape)
  require(dplyr)
  require(geiger)
  
  # format data frame input
  dd <- x %>% 
    as.tibble() %>% # coerce to tibble
    rename(taxon = taxa) %>% # identify taxon column
    filter(!is.na(taxon)) # remove NA taxa
  
  # specify all columns if none given by user
  if(is.null(cols)) cols <- names(dd %>% select(-taxon)) 
  
  # prune tree for matches only
  phy <- drop.tip(phy, setdiff(phy$tip.label, dd$taxon))
  phy <- multi2di(phy, random=FALSE) 
  phy$node.label <- paste("node", 
                          formatC(c(1:phy$Nnode), 
                                  digits=0, width=nchar(phy$Nnode), flag="0"))
  
  # get single PICs function
  pic1 <- function(col  # 2 column data frame or tibble with first column taxon, 
                   # second column trait data, possibly with NAs
                   ,  phy
                   ,  scale=scale) {  # a .tre object to patch first column of col

    # remove NA and add taxon as rownames for apre::pic 
    col <- col %>% 
      filter(!is.na(.[2])) %>%
      as.data.frame()
    rownames(col) <- col$taxon
    col <- col %>% 
      select(2)
    
    # prune data and tree
    col <- col[rownames(col) %in% intersect(rownames(col), phy$tip.label), 
               , drop=FALSE] 
    pruned <- drop.tip(phy, setdiff(phy$tip.label, rownames(col))) %>%
      compute.brlen()
    
# loop through lambda's for Pagel transformations
    corrs <- vector()
    modpics <- list()
    pvs <- vector()
    trans=vector()
    params=vector()
    
    for (t in c(
                # "allequal", 
                # "grafen",
                "pagel")) {
    
    if(t == "pagel") parms <- seq(0, 1, by=.1) else
      if (t == "grafen") parms <- c(seq(.1, 1, by=.1), seq(1.5, 10, by=.5)) else
        if (t == "allequal") parms <- 1
      
    for (p in parms) {
      
      if(t == "pagel") t.phy <-rescale(x=pruned, model="lambda", p) else
      if (t == "grafen") t.phy <- compute.brlen(phy=pruned, power=p) else
      if (t == "allequal") t.phy <- compute.brlen(phy=pruned, runif, min=p, max=p)
    # get pic column
    pic.out <- as.data.frame(pic(col[[1]], t.phy, scaled=scale, var.contrasts=TRUE))
    
    corr <- cor(abs(pic.out[,1]), sqrt(pic.out[,2]))
    test <- cor.test(abs(pic.out[,1]), sqrt(pic.out[,2]))
    
    ggplot() +
      geom_point(aes(x=sqrt(pic.out[,2]), y=abs(pic.out[,1])))
    
    corrs <- c(corrs, abs(corr))
    modpics <- c(modpics, list(pic.out))
    pvs <- c(pvs, test$p.value)
    trans <- c(trans, t)
    params <- c(params, p)
    
    }
  }
    
    pic.out.l <- list(pics = modpics[corrs == min(corrs)][[1]], 
                      trans = trans[corrs == min(corrs)],
                      parm=params[corrs == min(corrs)],
                      corr = min(corrs),
                      pvalue = pvs[corrs == min(corrs)])
    return(pic.out.l)
    
    }  
    
  
  # generate list of pics for all traits 
  pic_list <- lapply(cols, function(y) {
    
    trait <- data.frame(taxon=dd[["taxon"]], dd[[y]])
    names(trait)[2] <- y
    pic1(col=trait, phy=phy, scale=scale)
    
  })                     
  
  ojoin_list <- function(right, left) { # recursive joining function
    
    left <- right[1]
    right <- right[-1]
    
    if(length(right) == 1) return(full_join(left[[1]], right[[1]], by="node"))
    
    full_join(left[[1]], ojoin_list(right), by="node")
    
  }
  
  pics_only <- lapply(pic_list, function(x) data.frame(node=rownames(x$pics), contrasts=x$pics$contrasts))
  vars_only <- lapply(pic_list, function(x) data.frame(node=rownames(x$pics), variance=x$pics$variance))

  
  # join list of pics into single data frame, joining on node
  
  if(center) {pics_df <- suppressWarnings(ojoin_list(pics_only)) %>% 
    mutate_at(vars(2:ncol(.)), funs(. - mean(., na.rm=TRUE)))} else
      pics_df <- ojoin_list(pic_list)
  
  vars_df <- suppressWarnings(ojoin_list(vars_only))
  
  names(pics_df) <- c("node", cols)
  names(vars_df) <- c("node", cols)
  
  outlist <- list(pics_df=pics_df[order(pics_df$node), ], 
                  vars_df=vars_df[order(vars_df$node), ],
                  parms = data.frame(trait=cols,
                                     trans=sapply(pic_list, function(x) x$trans),
                                     parms=sapply(pic_list, function(x) x$parm),
                                     corrs=sapply(pic_list, function(x) x$corr),
                                     pvs = sapply(pic_list, function(x) x$pvalue)))
  
  return(outlist)
  
}
```

```{r PICs Function Testing}

## pic1 function

# specify variables
col <- sp.raw %>% select(taxon.new, cn)
phy <- my_phy <- read.tree(file = "~/Dropbox/Projects/Sev/Plants/Plant Phylogenies/sev_traits_tree101.tre")


pic1 <- function(col, phy) {
  
  phy$node.label <- paste("node", 1:phy$Nnode)
  
  # remove NA but keep rownames
  col <- col %>% 
    filter(!is.na(.[2])) %>%
    as.data.frame()
  rownames(col) <- col$taxon
  col <- col %>% 
    select(2)
  
  # prune data and tree
  col <- col[rownames(col) %in% intersect(rownames(col), phy$tip.label), , drop=FALSE] 
  pruned <- drop.tip(phy, setdiff(phy$tip.label, rownames(col)))
  
  # if(!setequal(pruned$tip.label, rownames(col))) # check that taxa match
  # return("ERROR: tree and trait taxa do not match")
  
  # resolve multifurcations with branches of zero length
  pruned <- multi2di(pruned, random=FALSE) 
  
  # get pic column
  pic.out <- pic(col[[1]], pruned) %>%
    tibble(names(.)) %>%
    select(2, 1)
  names(pic.out) <- c("node", names(col))
  
  return(pic.out)
  
} 

## make list of pics to feed to pic1

pic_list <- lapply(cols, function(x) {
  
  trait <- data.frame(taxon=dd[["taxon"]], dd[[x]])
  names(trait)[2] <- x
  pic1(col=trait, phy=my_phy)
  
})   
## whole function
cols

pics_out <- pics(x = sp.raw, cols=cols, taxa="taxon.new", phy=my_phy, center=TRUE)
```

```{r Create PICs}
library(geiger)

my_phy <- read.tree(file = "~/Dropbox/Projects/Sev/Plants/Plant Phylogenies/sev_traits_tree101.tre") 

# no transform
phy_in <- my_phy
  thepics <- pics(x = all_traits, cols=c("a_p") , phy=phy_in, taxa="taxon.new", scale=TRUE)
  print(ggplot() + 
    geom_point(aes(x=sqrt(thepics$vars_df[,trait]), y=abs(thepics$pics_df[,trait]))))

# grafen
rho=1
phy_in <- compute.brlen(my_phy, power=rho)
  thepics <- pics(x = all_traits, cols=c("a_p", "ldmc") , phy=phy_in, taxa="taxon.new", scale=TRUE)
  print(ggplot() + 
    geom_point(aes(x=sqrt(thepics$vars_df[,trait]), y=abs(thepics$pics_df[,trait]))))
 
# lambda (Pagel?) 
lambda=0
phy_in <- rescale(my_phy, "lambda", lambda)
  thepics <- pics(x = all_traits, cols=c("a_p", "ldmc") , phy=phy_in, taxa="taxon.new", scale=TRUE)
  print(ggplot() + 
    geom_point(aes(x=sqrt(thepics$vars_df[,trait]), y=abs(thepics$pics_df[,trait]))))
```

```{r}

my_phy <- read.tree(file = "~/Dropbox/Projects/Sev/Plants/Plant Phylogenies/sev_traits_tree101.tre") %>%
  compute.brlen(1)
my_phy$edge.length


# try different functions
for(trait in c("a_p", "ldmc", "lma", "cn")) {


parms_df <- do.call(rbind, pic_list <- lapply(seq(0.05, 1, by=0.05), function(rho) {
  
    phy_in <- compute.brlen(my_phy, "Grafen", power=rho)
    # phy_in <- compute.brlen(my_phy, 1)
  thepics <- pics(x = all_traits, cols=trait, phy=phy_in, taxa="taxon.new", scale=TRUE)
  data.frame(transform="grafen", param=rho, node=thepics$pics_df[[1]], 
             pics=thepics$pics_df[[2]], variance=thepics$vars_df[[2]])
  
}))

pdf_path <- paste("~/Dropbox/Projects/Sev/Plants/Figures/", 
                  trait, "_brlen.pdf", sep="")

pdf(file = pdf_path)

tplot <- parms_df %>%
  as.tibble() %>%
ggplot(aes(x=sqrt(variance), y=abs(pics))) + 
    geom_point() + 
    geom_smooth(method = "loess", colour = "red", fill = "red") +
    facet_wrap(~ param, scales="free")

print(tplot)

dev.off()

}



# piccorrs <- do.call(rbind,lapply(pic_list, function(df) {
#   data.frame(param=max(df$param), cor=cor(x=sqrt(df$variance), 
#                                               y=abs(df$pics)))}))
```


```{r Evol Models}
## test evolutionary models
my_phy <- read.tree(file = "~/Dropbox/Projects/Sev/Plants/Plant Phylogenies/sev_traits_tree101.tre") %>%
  compute.brlen() %>% multi2di(random=FALSE) 

lambdaopt <- vector()
for (t in names(all_pics$pics_df[2:ncol(all_pics$pics_df)])) {
  
ldmcdat <- as.data.frame(all_traits[t])
rownames(ldmcdat) <- all_traits$taxon.new
lambdatest <- fitContinuous(phy=my_phy, dat=ldmcdat, model="lambda")
lambdaopt <- c(lambdaopt, lambdatest$opt$lambda)

}

deltaopt <- vector()
for (t in names(all_pics$pics_df[2:ncol(all_pics$pics_df)])) {
  
ldmcdat <- as.data.frame(all_traits[t])
rownames(ldmcdat) <- all_traits$taxon.new
deltatest <- fitContinuous(phy=my_phy, dat=ldmcdat, model="delta")
deltaopt <- c(deltaopt, deltatest$opt$delta)

}

kappaopt <- vector()
for (t in names(all_pics$pics_df[2:ncol(all_pics$pics_df)])) {
  
ldmcdat <- as.data.frame(all_traits[t])
rownames(ldmcdat) <- all_traits$taxon.new
kappatest <- fitContinuous(phy=my_phy, dat=ldmcdat, model="OU")
kappaopt <- c(kappaopt,kappatest$opt$kappa)

}

ouopt <- vector()
for (t in names(all_pics$pics_df[2:ncol(all_pics$pics_df)])) {
  
ldmcdat <- as.data.frame(all_traits[t])
rownames(ldmcdat) <- all_traits$taxon.new
outest <- fitContinuous(phy=my_phy, dat=ldmcdat, model="BM")
ouopt <- c(ouopt,kappatest$opt$kappa)

}

```



##### PCAs
```{r PCA on PICs}

ccd <- all_pics.noscale %>%
  select(node, a_p, p95.height_A, ldmc:seed_mass, -sla) %>%
  filter(complete.cases(.)) 

pca.out <- ccd %>%
  select(-node) %>%
  prcomp(scale=TRUE)

plot(summary(pca.out)$importance[3,])
biplot(pca.out, choices=c(1, 2))

# combine with cv data
pca.all <- ccd %>% 
  select(node) %>%
  cbind(pca.out$x) %>%
  as.tibble() %>%
  inner_join(all_pics.noscale, by="node")
```

```{r Analysis with PICs}

ggplot(pca.all, aes(x=seed_mass)) +
  geom_histogram()

pca.all %>% filter( seed_mass > .002)
n3 <- extract.clade(phy, node="node 70")
plot.phylo(n3, show.node.label = TRUE)

pca.all %>% filter(!node %in% c("node 55", "node 70")) %>%
ggplot(aes(y=F.A.cv_BIM, x=PC3, label=node)) +
  geom_point() + 
  geom_label(label.size = .01)

an.PC <- lm(F.A.cv_BM ~ PC1 + PC2 + PC3, data=pca.all, subset=!node %in% c("node 55", "node 70"))
summary(an.PC)

an.a_p <- lm(F.A.cv_BM ~ a_p, data=pca.all)
summary(an.a_p)

an.ldmc <- lm(F.A.cv_BM ~ ldmc, data=pca.all)
summary(an.ldmc)

an.sdens <- lm(F.A.cv_BM ~ sdens, data=pca.all)
summary(an.sdens)

an.comb <- lm(F.A.cv_BM ~ ldmc + cn + fr.diam + sdens + seed_mass + a_p, data=pca.all)
summary(an.comb)
```




